---
title: "Admixture_for_students"
author: "Oscar Lao"
date: "2026-01-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(ggplot2)
library(slendr)
library(tibble)
library(dplyr)
library(BEDMatrix);
library(LEA)
library(genio);
library(smacof);
Sys.setenv(PATH = paste("C:/msys64/ucrt64/bin", Sys.getenv("PATH"), sep=";"))
init_env()
rm(list=ls());
```


#########################################################################
# 1. Project in two dimensions the relationships of a dataset generated with the basic model
# Goal: to understand how PCA, MDS and SMACOF works in practice.
# 
#


```{r}
# In Slendr and msprime
rm(list=ls());

rec_rate <- 0 # uniform recombination rate
mut_rate <- 1e-7 # mutation rate

basic_model <- function(time_split_A_B, Ne)
{
  # African ancestral population
  popA <- population("popA", time = 2*time_split_A_B, N = Ne)
  popB <- population("popB", parent = popA, time = time_split_A_B, N = Ne)
  model_nogf <- compile_model(populations = list(popA, popB), generation_time = 1,  simulation_length = 2*time_split_A_B)
  samples <- schedule_sampling(
  model_nogf, times = 0,
  list(popA, 50), list(popB, 50)
)
  return(model_nogf);
}


make_sim <- function(model_nogf)
{
  popA <- model_nogf$populations[[1]]
  popB <- model_nogf$populations[[2]]
  samples <- schedule_sampling(
  model_nogf, times = 0,
  list(popA, 50), list(popB, 50)
)  
  # number of independent genomic fragments
  n_fragments <- 500
  ts_list <- vector("list", n_fragments)
  for (i in 1:n_fragments) {
    ts_list[[i]] <- msprime(
    model_nogf,
    sequence_length = 1e3,   # length per fragment
    recombination_rate = rec_rate,
    samples = samples
    )
    ts_list[[i]] <- ts_mutate(ts_list[[i]], mut_rate)
  }

# genotypes
  geno_list <- lapply(ts_list, function(ts) {
  t(ts$genotype_matrix())   # SNPs × samples
})
  geno_all <- do.call(cbind, geno_list);
  geno_all_filtered <- geno_all[,apply(geno_all,2,var) > 0]

inds <- ts_list[[1]]$tables$individuals
nodes <- ts_list[[1]]$tables$nodes

# Get mapping: node → individual
node_ind <- nodes$individual

geno <- geno_all_filtered

# group haplotypes by individual
diploid_matrix <- sapply(unique(node_ind[node_ind != -1]), function(ind_id) {
  haplos <- which(node_ind == ind_id)
  colSums(geno[haplos, , drop = FALSE])
})

diploid_matrix <- t(diploid_matrix)
  
return(diploid_matrix);  
}

time_split_A_B <- 10000;
Ne <- 10000;

model <- basic_model(time_split_A_B, Ne);

plot_model(model);

diploid_matrix <- make_sim(model);

pca <- prcomp(diploid_matrix, scale. = TRUE)

df <- as.data.frame(pca$x[,1:2])
colnames(df) <- c("PC1", "PC2")
df$population <- rep(c("popA","popB"), each=50)

ggplot(df, aes(x = PC1, y = PC2, color = population)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "PCA of Simulated Genotypes",
    x = "PC 1",
    y = "PC 2"
  )

# MDS

# create a distance matrix between pairs of individuals. Distance can be whatever you want
# Euclidean
md <- dist(diploid_matrix)

mds.res <- cmdscale(md);

df <- as.data.frame(mds.res)
colnames(df) <- c("DIM1", "DIM2")
df$population <- rep(c("popA","popB"), each=50)

ggplot(df, aes(x = DIM1, y = DIM2, color = population)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "MDS of Simulated Genotypes",
    x = "PC 1",
    y = "PC 2"
  )

# SMACOF. Check ?type and run it with different configurations. What do you get?

smac <- mds(md, type = "ordinal")

df <- as.data.frame(smac$conf)
colnames(df) <- c("DIM1", "DIM2")
df$population <- rep(c("popA","popB"), each=50)

ggplot(df, aes(x = DIM1, y = DIM2, color = population)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "SMACOF of Simulated Genotypes",
    x = "PC 1",
    y = "PC 2"
  )

```
# run the algorithms with different configurations. Imagine I want to know the eigenvalues of the classical mds. What would you do? Can I estimate these values with SMACOF? Why? What would I provide instead of that?


# Exercise. Check the code and try to understand what is it doing. Modify the time of split and the effective population size. What happens when the time of split increases? What if the Ne decreases?

```{r}
# In Slendr and msprime
rm(list=ls());

rec_rate <- 0 # uniform recombination rate
mut_rate <- 1e-7 # mutation rate

basic_model_Ne1_Ne2 <- function(time_split_A_B, Ne1, Ne2, sample1, sample2)
{
  # TAKE AS TEMPLATE THE PREVIOUS CODE AND MODIFY IT TO ADD THE NEW PARAMETERS
  # YOUR CODE
return(diploid_matrix);  
}

time_split_A_B <- 10000;
Ne1 <- 1000;
Ne2 <- 10000;
sample1 <- 50;
sample2 <- 50;

model <- basic_model_Ne1_Ne2(time_split_A_B, Ne1, Ne2, sample1, sample2);

plot_model(model);

diploid_matrix <- make_sim(model);

pca <- prcomp(diploid_matrix, scale. = TRUE)

df <- as.data.frame(pca$x[,1:2])
colnames(df) <- c("PC1", "PC2")
df$population <- c(rep("popA",sample1),rep("popB",sample2))

ggplot(df, aes(x = PC1, y = PC2, color = population)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "PCA of Simulated Genotypes",
    x = "PC 1",
    y = "PC 2"
  )
```


#########################################################################
#
# 1.3. What happens if the samples come from DIFFERENT time transects?
# Change the time of split, the number of sampled points by time, etc.
#

```{r}
rm(list=ls());
rec_rate <- 0
mut_rate <- 1e-7

basic_model <- function(time_split_A_B, NeA, NeB,
                        sampleA, sampleB,
                        times,
                        simulation_length = 100000) {
  # ancestral population
  popA <- population("popA", time = simulation_length, N = NeA)

  # descendant population splits from popA at time_split_A_B
  popB <- population("popB", parent = popA, time = time_split_A_B, N = NeB)

  model <- compile_model(
    populations = list(popA, popB),
    generation_time = 1,
    simulation_length = simulation_length
  )

  # schedule time-series sampling
  samples <- schedule_sampling(
    model,
    times = times,
    list(popA, sampleA),
    list(popB, sampleB),
    strict = TRUE
  )

  list(model = model, samples = samples)
}

make_sim <- function(model, samples,
                     n_fragments = 100,
                     sequence_length = 1e3) {

  ts_list <- vector("list", n_fragments)

  for (i in seq_len(n_fragments)) {
    ts <- msprime(
      model,
      sequence_length = sequence_length,
      recombination_rate = rec_rate,
      samples = samples
    )
    ts <- ts_mutate(ts, mut_rate)
    ts_list[[i]] <- ts
  }

  # build haploid genotype matrix across fragments
  geno_list <- lapply(ts_list, function(ts) {
    t(ts$genotype_matrix())   # rows = haplotypes, cols = SNPs
  })

  geno_all <- do.call(cbind, geno_list)

  # keep only polymorphic SNPs
  geno_all_filtered <- geno_all[, apply(geno_all, 2, var) > 0, drop = FALSE]

  if (ncol(geno_all_filtered) == 0) {
    stop("No polymorphic SNPs were generated. Try increasing mut_rate, sequence_length, or n_fragments.")
  }

  # recover sample metadata from first tree sequence
  ts0 <- ts_list[[1]]

  node_info <- ts_nodes(ts0, sf = FALSE) |>
    tibble::as_tibble()

  sample_ind_ids <- node_info$ind_id[node_info$sampled]
  ordered_ind_ids <- unique(sample_ind_ids)

  # collapse haplotypes into diploid genotypes
  diploid_matrix <- t(sapply(ordered_ind_ids, function(ind) {
    haplos <- which(sample_ind_ids == ind)

    if (length(haplos) != 2) {
      stop(paste("Individual", ind, "has", length(haplos), "haplotypes instead of 2"))
    }

    colSums(geno_all_filtered[haplos, , drop = FALSE])
  }))

  sample_info <- node_info |>
    dplyr::filter(sampled) |>
    dplyr::distinct(ind_id, name, pop, time) |>
    dplyr::filter(ind_id %in% ordered_ind_ids) |>
    dplyr::slice(match(ordered_ind_ids, ind_id))

  list(
    diploid_matrix = diploid_matrix,
    sample_info = sample_info,
    ts_list = ts_list
  )
}

time_split_A_B <- 10000
NeA <- 1000
NeB <- 1000
sampleA <- 1
sampleB <- 1
times <- seq(from = 1, to = time_split_A_B -1, by = 100)

sim_setup <- basic_model(
  time_split_A_B = time_split_A_B,
  NeA = NeA,
  NeB = NeB,
  sampleA = sampleA,
  sampleB = sampleB,
  times = times,
  simulation_length = time_split_A_B*2
)

model <- sim_setup$model
samples <- sim_setup$samples

plot_model(model)

sim_result <- make_sim(model, samples)

diploid_matrix <- sim_result$diploid_matrix
sample_info <- sim_result$sample_info
ts_list <- sim_result$ts_list

pca <- prcomp(diploid_matrix, scale. = TRUE)

df <- as.data.frame(pca$x[, 1:2])
colnames(df) <- c("PC1", "PC2")
df$population <- sample_info$pop
df$time <- sample_info$time
df$name <- sample_info$name

ggplot(df, aes(x = PC1, y = PC2, color = population)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "PCA of Simulated Genotypes",
    x = "PC1",
    y = "PC2"
  )
```

# Why are we observing this shape? What does that mean? Change the time of split to older and closer to present times. What do you see? Discuss it with the other students. Who gets the more strange shape?


#########################################################################
# 2. Project in two dimensions the relationships of a dataset generated with the model.admixed.pop
# Goal: to understand how PCA works in practice with admixed populations
# Exercise. Check what happens when the time of admixture increases. What if the proportion of admixture changes?


```{r}
rm(list=ls());

rec_rate <- 0 # uniform recombination rate
mut_rate <- 1e-7 # mutation rate


basic_admix_model <- function(time_split_A_B, time_admix, Ne, admix_B)
{
  # African ancestral population
  popA <- population("popA", time = 2*time_split_A_B, N = Ne)
  popB <- population("popB", parent = popA, time = time_split_A_B, N = Ne)
  popAdmix <- population("popAdmix", parent = popA, time = time_admix, N = Ne)
  gf <- gene_flow(from = popB, to = popAdmix, start = (time_admix-1), end = (time_admix-2), proportion = admix_B)
  model_nogf <- compile_model(populations = list(popA, popB, popAdmix), generation_time = 1,   gene_flow = gf, direction = "backward",
 simulation_length = 2*time_split_A_B)
  return(model_nogf);
}


make_sim <- function(model_nogf)
{
  popA <- model_nogf$populations[[1]]
  popB <- model_nogf$populations[[2]]
  popAdmix <- model_nogf$populations[[3]]
  samples <- schedule_sampling(
  model_nogf, times = 0,
  list(popA, 50), list(popB, 50), list(popAdmix, 50)
)  
  # number of independent genomic fragments
  n_fragments <- 100
  ts_list <- vector("list", n_fragments)
  for (i in 1:n_fragments) {
    ts_list[[i]] <- msprime(
    model_nogf,
    sequence_length = 1e3,   # length per fragment
    recombination_rate = rec_rate,
    samples = samples
    )
    ts_list[[i]] <- ts_mutate(ts_list[[i]], mut_rate)
  }

# genotypes
  geno_list <- lapply(ts_list, function(ts) {
  t(ts$genotype_matrix())   # SNPs × samples
})
  geno_all <- do.call(cbind, geno_list);
  geno_all_filtered <- geno_all[,apply(geno_all,2,var) > 0]
  

inds <- ts_list[[1]]$tables$individuals
nodes <- ts_list[[1]]$tables$nodes

# Get mapping: node → individual
node_ind <- nodes$individual

geno <- geno_all_filtered

# group haplotypes by individual
diploid_matrix <- sapply(unique(node_ind[node_ind != -1]), function(ind_id) {
  haplos <- which(node_ind == ind_id)
  colSums(geno[haplos, , drop = FALSE])
})

diploid_matrix <- t(diploid_matrix)
  
return(diploid_matrix);  
}

time_split_A_B <- 5000;
time_admix <- 10;
Ne <- 100;
admix_B <- 0.1;

demographic_model <- basic_admix_model(time_split_A_B, time_admix, Ne, admix_B);

plot_model(demographic_model);
  
diploid_matrix <- make_sim(demographic_model);

pca <- prcomp(diploid_matrix, scale. = TRUE)

df <- as.data.frame(pca$x[,1:2])
colnames(df) <- c("PC1", "PC2")
df$population <- rep(c("popA","popAdmix","popB"), each=50)

ggplot(df, aes(x = PC1, y = PC2, color = population)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "PCA of Simulated Genotypes",
    x = "PC 1",
    y = "PC 2"
  )
```

rm(list=ls());

#########################################################################
# Pure admixture model defined with fastSimcoal2
#
#

# FasSimcoal2 is another coalescent-based simulator. It was originally developed by Laurent Excoffier.

```{r}
rm(list=ls());

# define the folder where the executable of fastSimcoal2 is stored
folder.fastSimcoal2 <- "C:\\Users\\u9424\\OneDrive - Universitat Pompeu Fabra\\Grants\\2026\\EMBO Naples 2026\\Second day\\Morning\\Practical\\";

# Executable
exe <- ".\\fsc28.exe"

setwd(folder.fastSimcoal2);

model.admixed.pop <- function(sample_1, sample_2, sample_admix, effective_population_size_1, effective_population_size_2, effective_population_size_admixed, time_admixture, percentage_from_population_1, time_split, effective_population_size_ancestral, number_of_blocks)
{
  lines <- c(
    "//Number of population samples (demes)",
    "3",
    "//Population effective sizes (number of genes)",
    effective_population_size_1,
    effective_population_size_2,
    effective_population_size_admixed,
    "//Sample sizes",
    sample_1,
    sample_admix,
    sample_2,
    "//Growth rates\t: negative growth implies population expansion",
    "0",
    "0",
    "0",
    "//Number of migration matrices : 0 implies no migration between demes",
    "0",
    "//historical event: time, source, sink, migrants, new size, new growth rate, migr. matrix",
    "3 historical event",
    paste(time_admixture,"1 0",percentage_from_population_1,"1 0 0",sep = " "),
    paste(time_admixture,"1 2 1 1 0 0",sep = " "),
    paste(time_split,"0 2 1",effective_population_size_ancestral/effective_population_size_2, "0 0",sep = " "),
    "//Number of independent loci [chromosome]",
    paste(number_of_blocks,"1",sep=" ")
  )
  
  # Repeating the block 22 times
  block <- c(
    "//Per chromosome: Number of linkage blocks",
    "1",
    "//per Block: data type, num loci, rec. rate and mut rate + optional parameters",
    "DNA 1000000 1.0E-8 1.855284327902964E-8 0.0"
  )
  
  # Append number_of_blocks blocks
  for (i in 1:number_of_blocks) {
    lines <- c(lines, block)
  }
  
  # Write to file
  writeLines(lines, "DemographicModelSplitR.par")
  
  args <- c("-i", ".\\DemographicModelSplitR.par", "-x", "-s0", "-d", "-n", "1", "-q", "-G")
  # Execute
  system2(exe, args = args)
  # Read the output
  data.t <- read.table(file=paste(folder.fastSimcoal2,"\\DemographicModelSplitR\\DemographicModelSplitR_1_1.gen", sep=""), header = T);
  
  # First four columns are snp info
  # haplotype matrix. Rows are haplotypes, columns are positions
  
  H <- t(as.matrix(data.t[,5:ncol(data.t)]));
  # First 100 haplotypes is pop 1, next pop 2
  rownames(H) <- c(rep("A",sample_1),rep("Admix",sample_admix),rep("B",sample_2));
  return(list(position = data.t[,1:4],haplotype_matrix = H));
}
```


#########################################################################
# 2.1. Project in two dimensions the relationships of a dataset generated with the model.admixed.pop generated with fastSimcoal2
# Goal: to understand how MDS works in practice with an admixed population
# Exercise:
# Create a dataset of 100 chromosomes in each population. The effective population size of pop1 is 1000, the same for pop 2 and pop 3.
# the time of admixture is set to 10 generations and the time of split between the two parental populations to 10000 generations ago.
# the effective population size ancestral is 2000.
# the percentage of admixture from population 1 is 0.3.
# run the model. Generate a distance matrix and run a classical multidimensional scaling (cmdscale).
# Plot the first two dimensions.

```{r}

sample_1 <- 100;
sample_2 <-100;
sample_admix <- 100;
effective_population_size_1 <- 1000;
effective_population_size_2 <- 1000;
effective_population_size_admixed <- 1000;
time_admixture <- 10;
time_split <- 10000;
effective_population_size_ancestral <- 2000;
number_of_blocks <- 10;
percentage_from_population_1 <- 0.3;
sim <- model.admixed.pop(sample_1, sample_2, sample_admix, effective_population_size_1, effective_population_size_2, effective_population_size_admixed, time_admixture, percentage_from_population_1, time_split, effective_population_size_ancestral, number_of_blocks)
# first 5 columns are information of the SNPs
haplotypes <- sim$haplotype_matrix;
# Genotypes are computed by collapsing two contiguous haplotypes
genotypes <- haplotypes[seq(from=1,to=nrow(haplotypes),by=2),] + haplotypes[seq(from=2,to=nrow(haplotypes),by=2),];
# Provide labels
pop <- rownames(haplotypes)[seq(from=1,to=nrow(haplotypes),by=2)];
rownames(genotypes) <- pop;
# Compute the standard deviation of each marker to decide if it is worth being included
sd.snps <- apply(genotypes,2,sd);
consider.snps <- which(sd.snps > 0.1);
genotypes.clean <- genotypes[,consider.snps];
# scale each marker by substracting the mean and dividing by the standard deviation
genotypes.clean.scaled <- scale(genotypes.clean);
# compute the euclidean distance between each two individuals
dmatrix <- dist(genotypes.clean.scaled);
# conduct the classical multidimensional scaling
mds.result <- cmdscale(dmatrix, k = (nrow(as.matrix(dmatrix))-1),add=T,eig=T);
points.mds <- mds.result$points[,1:2];
rownames(points.mds) <- pop;

df <- as.data.frame(mds.result$points[,1:2])
colnames(df) <- c("Dim1", "Dim2")
df$population <- rep(c("A","Admix","B"), each=50)

ggplot(df, aes(x = Dim1, y = Dim2, color = population)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "MDS of Simulated Genotypes",
    x = "Dim 1",
    y = "Dim 2"
  )
```

#########################################################################
# 2.2. Basic admixture estimator
# Goal: to understand how the dimensions of the MDS relate to admixture parameters
# Exercise:
# Create a function that takes the first dimension of the MDS eigenvector output.
# Rescales the values of the Admix individuals to (x-meanPop1)/(meanPop2-meanPop1)
# Comment each line of the code. What is the code doing?


```{r}
extract.admixture <- function(points.mds)
{
  popA <- which(rownames(points.mds)=="A");
  popB <- which(rownames(points.mds)=="B");
  popAd <- which(rownames(points.mds)=="Admix");
  meanA <- mean(points.mds[popA,1]);
  meanB <- mean(points.mds[popB,1]);
  return(1-(points.mds[popAd,1]-meanA)/(meanB-meanA));
}
```


#########################################################################
# 2.3 Admixture proportions and position in the first dimension of the scaled MDS
# Goal: to understand how the dimensions of the MDS relate to admixture parameters
# Exercise:
# Generate 1000 samples, changing at random the percentage of admixture from population 1

```{r}
sample_1 <- 100;
sample_2 <-100;
sample_admix <- 100;
effective_population_size_1 <- 1000;
effective_population_size_2 <- 1000;
effective_population_size_admixed <- 1000;
time_admixture <- 10;
time_split <- 10000;
effective_population_size_ancestral <- 2000;
number_of_blocks <- 10;

percentage.admixture <- rep(NA,100);
percentage.admixture.estimated <- rep(NA,100);
for(rep in 1:100)
{
  percentage_from_population_1 <- runif(1,0,1);
  percentage.admixture[rep] <- percentage_from_population_1;
  sim <- model.admixed.pop(sample_1, sample_2, sample_admix,effective_population_size_1, effective_population_size_2, effective_population_size_admixed, time_admixture, percentage_from_population_1, time_split, effective_population_size_ancestral, number_of_blocks)
  # first 5 columns are information of the SNPs
  haplotypes <- sim$haplotype_matrix;
  # Genotypes are computed by collapsing two contiguous haplotypes
  genotypes <- haplotypes[seq(from=1,to=nrow(haplotypes),by=2),] + haplotypes[seq(from=2,to=nrow(haplotypes),by=2),];
  # Provide labels
  pop <- rownames(haplotypes)[seq(from=1,to=nrow(haplotypes),by=2)];
  rownames(genotypes) <- pop;
  # Compute the standard deviation of each marker to decide if it is worth being included
  sd.snps <- apply(genotypes,2,sd);
  consider.snps <- which(sd.snps > 0.1);
  genotypes.clean <- genotypes[,consider.snps];
  # scale each marker by substracting the mean and dividing by the standard deviation
  genotypes.clean.scaled <- scale(genotypes.clean);
  # compute the euclidean distance between each two individuals
  dmatrix <- dist(genotypes.clean.scaled);
  # conduct the classical multidimensional scaling
  mds.result <- cmdscale(dmatrix, k = (nrow(as.matrix(dmatrix))-1),add=T,eig=T);
  points.mds <- mds.result$points[,1:2];
  rownames(points.mds) <- pop;
  percentage.admixture.estimated[rep] <- median(extract.admixture(points.mds))  
}

plot(percentage.admixture, percentage.admixture.estimated);
summary(lm(percentage.admixture.estimated ~ percentage.admixture));
```

# What does this mean? Could we interpret admixture proportions with regards to the position of the admixed population and the parental populations?

#########################################################################
# 2.4. Variance in the admixture proportions and time of admixture
# Goal: to understand how the dimensions of the MDS relate to time of admixture parameters
# Exercise:
# Generate 100 samples, with a fixed percentage of admixture from population 1 of 0.5
# change the time of admixture between 1 generation to 100 generations
# estimate the variance in the percentage of estimated admixture
# plot the time  of admixture vs the variance in the percentage of estimated admixture

```{r}
effective_population_size_1 <- 1000;
effective_population_size_2 <- 1000;
effective_population_size_admixed <- 1000;
time_split <- 10000;
effective_population_size_ancestral <- 2000;
number_of_blocks <- 10;
percentage_from_population_1 <- 0.5;

time.admixture <- rep(NA,100);
sd.percentage.admixture.estimated <- rep(NA,100);
for(rep in 1:100)
{
  time_admixture <- round(runif(1,1,5000));
  time.admixture[rep] <- time_admixture;
  sim <- model.admixed.pop(sample_1, sample_2, sample_admix,effective_population_size_1, effective_population_size_2, effective_population_size_admixed, time_admixture, percentage_from_population_1, time_split, effective_population_size_ancestral, number_of_blocks)
  # first 5 columns are information of the SNPs
  haplotypes <- sim$haplotype_matrix;
  # Genotypes are computed by collapsing two contiguous haplotypes
  genotypes <- haplotypes[seq(from=1,to=nrow(haplotypes),by=2),] + haplotypes[seq(from=2,to=nrow(haplotypes),by=2),];
  # Provide labels
  pop <- rownames(haplotypes)[seq(from=1,to=nrow(haplotypes),by=2)];
  rownames(genotypes) <- pop;
  # Compute the standard deviation of each marker to decide if it is worth being included
  sd.snps <- apply(genotypes,2,sd);
  consider.snps <- which(sd.snps > 0.1);
  genotypes.clean <- genotypes[,consider.snps];
  # scale each marker by substracting the mean and dividing by the standard deviation
  genotypes.clean.scaled <- scale(genotypes.clean);
  # compute the euclidean distance between each two individuals
  dmatrix <- dist(genotypes.clean.scaled);
  # conduct the classical multidimensional scaling
  mds.result <- cmdscale(dmatrix, k = (nrow(as.matrix(dmatrix))-1),add=T,eig=T);
  points.mds <- mds.result$points[,1:2];
  rownames(points.mds) <- pop;
  sd.percentage.admixture.estimated[rep] <- sd(extract.admixture(points.mds))  
}

plot(time.admixture, sd.percentage.admixture.estimated);
```

How do we interpret these results? Is the variance in ancestry proportions a proxy for estimating the time when admixture took place? Until when? What parameters of the model can affect the shape of the function?


#########################################################################
# 3. PCA output from a 1D stepping stone model
# Goal: to understand how PCA works in a 1D stepping stone
# Exercise:
# Generate 8 populations, each with Ne = 500 chromosomes
# sample 20 chromosomes by population. Run PCA and plot the result with ggplot.

```{r}
rm(list=ls());
rec_rate <- 0;
mut_rate <- 1e-7 # mutation rate
make_1d_stepping_stone <- function(
  k,
  N = 1000,
  m_percent = 1,
  simulation_length = 10000,
  split_time = simulation_length - 1,
  sample_size = 2,
  sample_times = 1
) {
  stopifnot(k >= 2)
  stopifnot(N > 0)
  stopifnot(m_percent >= 0 && m_percent <= 100)

  # Convert percent to per-generation migration rate
  m <- m_percent / 100

  # Dummy ancestral population
  anc <- population(
    "ANC",
    time = simulation_length,
    N = N
  )

  # Create k present-day demes, all splitting from the ancestor
  pops <- lapply(seq_len(k), function(i) {
    population(
      name = paste0("pop", i),
      parent = anc,
      time = split_time,
      N = N
    )
  })

  names(pops) <- paste0("pop", seq_len(k))

  # Symmetric migration only between adjacent demes:
  # pop1 <-> pop2 <-> pop3 <-> ... <-> popk
  gf <- list()
  for (i in seq_len(k - 1)) {
    gf[[length(gf) + 1]] <- gene_flow(
      from = pops[[i]],
      to = pops[[i + 1]],
      start = split_time-1,
      end = 1,
      migration_rate = m,
      overlap = FALSE
    )

    gf[[length(gf) + 1]] <- gene_flow(
      from = pops[[i + 1]],
      to = pops[[i]],
      start = split_time-1,
      end = 1,
      migration_rate = m,
      overlap = FALSE
    )
  }

  # Compile for simulation
  model <- compile_model(
    populations = c(list(anc), pops),
    gene_flow = gf,
    generation_time = 1,
    simulation_length = simulation_length,
    serialize = FALSE
  )

  # Build a sampling schedule
  sample_args <- lapply(pops, function(p) list(p, sample_size))

  samples <- do.call(
    schedule_sampling,
    c(
      list(
        model = model,
        times = sample_times,
        strict = TRUE
      ),
      sample_args
    )
  )

  list(
    model = model,
    samples = samples,
    pops = pops,
    gene_flow = gf
  )
}

sim <- make_1d_stepping_stone(
  k = 8,                 # number of demes
  N = 1000,              # deme size
  m_percent = 0.02,         # 2% migrants per generation
  simulation_length = 5000,
  sample_size = 20,       # sampled diploid individuals per deme
  sample_times = 1       # sample at the present
)

# Visualize the demographic model
plot_model(sim$model)

# Run msprime
  # number of independent genomic fragments
  n_fragments <- 100
  ts_list <- vector("list", n_fragments)
  for (i in 1:n_fragments) {
    ts_list[[i]] <- msprime(
    sim$model,
    sequence_length = 1e3,   # length per fragment
    recombination_rate = rec_rate,
    samples = sim$samples
    )
    ts_list[[i]] <- ts_mutate(ts_list[[i]], mut_rate)
  }

# genotypes
  geno_list <- lapply(ts_list, function(ts) {
  t(ts$genotype_matrix())   # SNPs × samples
})
  geno_all <- do.call(cbind, geno_list);
  geno_all_filtered <- geno_all[,apply(geno_all,2,var) > 0]
  

inds <- ts_list[[1]]$tables$individuals
nodes <- ts_list[[1]]$tables$nodes

# Get mapping: node → individual
node_ind <- nodes$individual

geno <- geno_all_filtered

# group haplotypes by individual
diploid_matrix <- sapply(unique(node_ind[node_ind != -1]), function(ind_id) {
  haplos <- which(node_ind == ind_id)
  colSums(geno[haplos, , drop = FALSE])
})

diploid_matrix <- t(diploid_matrix)


# get annotated node table
node_info <- ts_nodes(ts_list[[1]], sf = FALSE) |>
  tibble::as_tibble()

# sampled haplotypes -> individual IDs
sample_ind_ids <- node_info$ind_id[node_info$sampled]
ordered_ind_ids <- unique(sample_ind_ids)

# collapse haplotypes to diploid individuals in the same order
diploid_matrix <- t(sapply(ordered_ind_ids, function(ind) {
  haplos <- which(sample_ind_ids == ind)
  colSums(geno_all_filtered[haplos, , drop = FALSE])
}))

# get metadata for those diploid individuals in the same order
sample_info <- node_info |>
  dplyr::filter(sampled) |>
  dplyr::distinct(ind_id, name, pop, time) |>
  dplyr::filter(ind_id %in% ordered_ind_ids) |>
  dplyr::slice(match(ordered_ind_ids, ind_id))

# PCA
pca <- prcomp(diploid_matrix, scale. = TRUE)

# plotting data frame
df <- as.data.frame(pca$x[, 1:2])
colnames(df) <- c("PC1", "PC2")

df$population <- sample_info$pop
df$individual <- sample_info$name
df$time <- sample_info$time

ggplot(df, aes(x = PC1, y = PC2, color = population)) +
  geom_point(size = 2, alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "PCA of Simulated Genotypes",
    x = "PC 1",
    y = "PC 2"
  )
```


#########################################################################
# 3.1 Grid of 2D stepping stone pops
# Goal: to understand from a coalescence point of view what admixture is in spatial models
# Exercise:
# Create function called model.admixed.pop with a fastSimcoal2 model that 
# models a two dimensional grid of populations, connected to the geographically closest elements.

```{r}
rm(list=ls())
# define the folder where the executable of fastSimcoal2 is stored
folder.fastSimcoal2 <- "C:\\Users\\u9424\\OneDrive - Universitat Pompeu Fabra\\Grants\\2026\\EMBO Naples 2026\\Second day\\Morning\\Practical\\";
# Executable
exe <- ".\\fsc28.exe"

model.2D.grid.pop <- function(sample_sizes, nes, migration_rate, number_of_blocks)
{
  if(length(sample_sizes)!=length(nes))
  {
    print("The lenght of the vector of sample_size must be the same as Ne!");
    print(c(length(sample_sizes),length(nes)));
    return(NA);
  }
  # size of grid
  n <- length(sample_sizes)^0.5
  
  if(round(n)-n!=0)
  {
    print("Length of populations must be pow of an integer");
    return(NA);
  }
  total_pops <- length(sample_sizes);
  grow_rates <- rep(0, total_pops);
  migration_grid <- matrix(nrow=total_pops, ncol = total_pops, 0);
  # Isolation by distance
  for(po in 1:total_pops)
  {
    pop <- po -1;
    row_pop1 <- floor(pop/n);
    col_pop1 <- pop%%n;
    for(po2 in 1:total_pops)
    {
      pop2 <- po2 - 1;
      row_pop2 <- floor(pop2/n);
      col_pop2 <- pop2%%n;      
      d <- (row_pop1-row_pop2)^2 + (col_pop1-col_pop2)^2;
      if(d!=0 && d <= 2)
      {
        # receives in forward from r, c isolation migrants
        migration_grid[po,po2] <- 1;                
      }      
    }
  }
  # scale by column
  for(c in 1:total_pops)
  {
    migration_grid[,c] <- migration_rate*migration_grid[,c];
  }
  
  mmatrix <- apply(migration_grid, 1, function(row) paste(row, collapse = " "))
  
  lines <- c(
    "//Number of population samples (demes)",
    total_pops,
    "//Population effective sizes (number of genes)",
    nes,
    "//Sample sizes",
    sample_sizes,
    "//Growth rates\t: negative growth implies population expansion",
    grow_rates,
    "//Number of migration matrices : 0 implies no migration between demes",
    "1",
    "//migration matrix",
    mmatrix,
    "//historical event: time, source, sink, migrants, new size, new growth rate, migr. matrix",
    "0 historical event",
    "//Number of independent loci [chromosome]",
    paste(number_of_blocks,"1",sep=" ")
  )
  
  # Repeating the block 22 times
  block <- c(
    "//Per chromosome: Number of linkage blocks",
    "1",
    "//per Block: data type, num loci, rec. rate and mut rate + optional parameters",
    "DNA 1000000 1.0E-8 1.855284327902964E-8 0.0"
  )
  
  # Append number_of_blocks blocks
  for (i in 1:number_of_blocks) {
    lines <- c(lines, block)
  }
  
  # Write to file
  writeLines(lines, "DemographicModelSplitR.par")
  
  args <- c("-i", ".\\DemographicModelSplitR.par", "-x", "-s0", "-d", "-n", "1", "-q", "-G")
  # Execute
  system2(exe, args = args)
  
  data.t <- read.table(file=paste(folder.fastSimcoal2,"\\DemographicModelSplitR\\DemographicModelSplitR_1_1.gen", sep=""), header = T);
  
  # First four columns are snp info
  # haplotype matrix. Rows are haplotypes, columns are positions
  
  H <- t(as.matrix(data.t[,5:ncol(data.t)]));
  # First 100 haplotypes is pop 1, next pop 2
  lab <- c();
  for(s in 1:total_pops)
  {
    lab <- c(lab,c(rep(s,sample_sizes[s])));
  }
  rownames(H) <- lab
  return(list(position = data.t[,1:4],haplotype_matrix = H));
}
```


#########################################################################
# 3.2. MDS output from a 2D stepping stone model
# Goal: to understand how an MDS works in a 2D stepping stone
# Exercise:
# Generate 16 populations (a grid of 4 by 4 populations), each with Ne = 500 chromosomes
# sample 20 chromosomes by population. Run MDS and plot the result with ggplot.

```{r}
sample_sizes <- rep(20,16);
Ne <- rep(500, 16);
migration_rate <- 10^-5;
number_of_blocks <- 10;
sim <- model.2D.grid.pop(sample_sizes, Ne, migration_rate, number_of_blocks)
# first 5 columns are information of the SNPs
haplotypes <- sim$haplotype_matrix;
# Genotypes are computed by collapsing two contiguous haplotypes
genotypes <- haplotypes[seq(from=1,to=nrow(haplotypes),by=2),] + haplotypes[seq(from=2,to=nrow(haplotypes),by=2),];
# Provide labels
pop <- rownames(haplotypes)[seq(from=1,to=nrow(haplotypes),by=2)];
rownames(genotypes) <- pop;
# Compute the standard deviation of each marker to decide if it is worth being included
sd.snps <- apply(genotypes,2,sd);
consider.snps <- which(sd.snps > 0.1);
genotypes.clean <- genotypes[,consider.snps];
# scale each marker by substracting the mean and dividing by the standard deviation
genotypes.clean.scaled <- scale(genotypes.clean);
# compute the euclidean distance between each two individuals
dmatrix <- dist(genotypes.clean.scaled);
# conduct the classical multidimensional scaling
mds.result <- cmdscale(dmatrix, k = (nrow(as.matrix(dmatrix))-1),add=T,eig=T);
points.mds <- mds.result$points[,1:2];
rownames(points.mds) <- pop;

ma <- data.frame(model = pop, x = points.mds[,1], y = points.mds[,2]);

ggplot(ma, aes(x=x, y=y, color=model)) +
  geom_point() +
  labs(
    x = paste("DIM 1 (",round(100*mds.result$eig[1]/sum(mds.result$eig),2) ," %)"),
    y = paste("DIM 2 (",round(100*mds.result$eig[2]/sum(mds.result$eig),2) ," %)")
  )
```


#########################################################################
# 4. Making a plink file
# Goal: to understand how to manage plink files
# Exercise:
# Create a dataset with the admixed model and store it in plink format

# Plink is a software for working with genotypic data (mainly GWAS). For population genomics, it is interesting because it has a quite efficient way of storing large matrices of genotypic data. Furthermore, some sofware also uses this kind of file format. For example, ADMIXTURE.

```{r}

# First, execute the piece of code that contains the admixed model from exercise 2.1. This will clear the environment from other models in order to prevent messing up

# Now we will generate a dataset with this model
sample_1 <- 100;
sample_2 <- 100;
sample_admix <- 100;
effective_population_size_1 <- 1000;
effective_population_size_2 <- 1000;
effective_population_size_admixed <- 1000;
time_split <- 10000;
effective_population_size_ancestral <- 2000;
number_of_blocks <- 10;
percentage_from_population_1 <- 0.5;
time.admixture <- rep(NA,100);
sd.percentage.admixture.estimated <- rep(NA,100);
time_admixture <- 20;
sim <- model.admixed.pop(sample_1, sample_2, sample_admix,effective_population_size_1, effective_population_size_2, effective_population_size_admixed, time_admixture, percentage_from_population_1, time_split, effective_population_size_ancestral, number_of_blocks)
# first 5 columns are information of the SNPs
haplotypes <- sim$haplotype_matrix;
# Genotypes are computed by collapsing two contiguous haplotypes
genotypes <- haplotypes[seq(from=1,to=nrow(haplotypes),by=2),] + haplotypes[seq(from=2,to=nrow(haplotypes),by=2),];
# Provide labels
pop <- rownames(haplotypes)[seq(from=1,to=nrow(haplotypes),by=2)];
rownames(genotypes) <- pop;
# Compute the standard deviation of each marker to decide if it is worth being included
sd.snps <- apply(genotypes,2,sd);
consider.snps <- which(sd.snps > 0.1);
genotypes.clean <- genotypes[,consider.snps];

# store the genotype matrix in Plink format
bim.data <- data.frame(sim$position[consider.snps,1], paste("rs",sim$position[consider.snps,2],sep=""), rep(0,length(consider.snps)), sim$position[consider.snps,2], sim$position[consider.snps,3],sim$position[consider.snps,4]);
colnames(bim.data) <- c("chr", "id", "posg", "pos", "alt", "ref");
fam.data <- data.frame(pop,paste(pop,seq(from=1,to=length(pop),by=1),sep="_"),rep(0,length(pop)),rep(0,length(pop)),rep(0,length(pop)),rep(-9,length(pop)));
colnames(fam.data) <- c("fam", "id", "pat", "mat", "sex", "pheno");
bed.matrix <- t(genotypes.clean);
colnames(bed.matrix) <- fam.data$id;
write_plink("modelAdmixture_1_1", bed.matrix, bim = bim.data, fam = fam.data);

# Lets check that the dataset was properly written
kk <- BEDMatrix("modelAdmixture_1_1")
dim(kk)
# You can use the bed file to run ADMIXTURE in linux
# Two files are produced. Q -> The admixture proportions of each individual. P -> the frequencies in the ancestral
```


#########################################################################
# 5. Estimating ancestry proportions with sNMF
# Goal: to estimate ancestry proportions using one algorithm for global ancestry estimation 
# Exercise:
# use sNMF, which has been coded in R, to estimate ancestry proportions. BEWARE: Results can be different
# from the ones from ADMIXTURE as the algorithm for ancestry optimization and the assumptions of the unsupervised model
# ARE DIFFERENT

# Step 1
# create the geno data
#genotypes.clean has n individuals as rows and m SNVs as columns
# we use the clean, because we will have to take into account 

```{r}
write.geno(genotypes.clean, "genotypes.geno")
project = NULL;
project = snmf("genotypes.geno",
               K = 1:5,
               entropy = TRUE,
               repetitions = 2,
               project = "new");

# How good is each model (at each K)?
plot(project, col = "blue", pch = 19, cex = 1.2)
# Extract Q matrix (ancestry coefficients)
ce <- cross.entropy(project, K = 2)
# Find the run with minimum cross-entropy
best <- which.min(ce)
Q <- as.matrix(Q(project, K = 2, run = best))
# Create an index that orders individuals by population label
ordered_indices <- order(pop)
# Reorder Q matrix and population labels
Q_ordered <- Q[ordered_indices, ]
pop_ordered <- pop[ordered_indices]
# Define colors
my.colors <- c("tomato", "gold")

# Plot the barplot manually using barplot(), not barchart()
barplot(t(Q_ordered),
        col = my.colors,
        border = NA,
        space = 0,
        xlab = "Individuals",
        ylab = "Ancestry proportions",
        main = "Ancestry matrix")

# Add population labels on the x-axis
axis(1, at = seq(0.5, nrow(Q_ordered) - 0.5, by = 1),
     labels = pop_ordered,
     las = 2,
     cex.axis = 0.4)
```
#Admixture proportions estimated by this way can also be used to estimate the time when such admixture happened and the amount of admixture. Clearly, as we have seen before with PCA and MDS, the SAME admixture porportions can be produced by DIFFERENT EVOLUTIONARY MODELS.

#KEY MESSAGE: These analyses are mainly EXPLORATORY.Using them as hypothesis testing only makes sense if we have a PRIOR knowledge about the demographic model that shaped the genetic variation of the samples.

#Remember: these algorithms are sensible to unequal sample size, presence of highly related individuals, different taxes of recombination across the genome, etc..


#########################################################################
# 6. Cluster Individuals after feature extraction
# Goal: Use clustering methods to define subgroups.
# Exercise:
# Run mclust and dbscan to visualize population clusters in the PCA.

```{r}
library(mclust);

rm(list= ls());
data.fibla <- read.table(file = "Location_lat_lon_population395.txt", header = T);

ibd.matrix <- matrix(nrow = nrow(data.fibla), ncol = nrow(data.fibla),0);
rownames(ibd.matrix) <- data.fibla$DNA_code;
colnames(ibd.matrix) <- data.fibla$DNA_code;

for(chr in 1:22)
{
  data.hbd <- read.table(gzfile(paste("out.chr", chr, ".hbd.gz", sep="")), header =F);
  for(e in 1:nrow(data.hbd))
  {
    name <- strsplit(data.hbd[e,1],"_")[[1]][2]
    r <- which(data.fibla$DNA_code==name);
    ibd.matrix[r,r] <- ibd.matrix[r,r] + data.hbd[e,8];
  }
  data.ibd <- read.table(gzfile(paste("out.chr", chr, ".ibd.gz", sep="")), header =F);
  for(e in 1:nrow(data.ibd))
  {
    name_one <- strsplit(data.ibd[e,1],"_")[[1]][2];
    name_two <- strsplit(data.ibd[e,3],"_")[[1]][2];
    r_one <- which(data.fibla$DNA_code==name_one);
    r_two <- which(data.fibla$DNA_code==name_two);
    ibd.matrix[r_one,r_two] <- ibd.matrix[r_one,r_two] + data.ibd[e,8];
    ibd.matrix[r_two,r_one] <- ibd.matrix[r_one,r_two];
  }  
}

dis.ibd.matri<- (max(ibd.matrix)- ibd.matrix)/(max(ibd.matrix)-min(ibd.matrix));

diag(dis.ibd.matri) <- 0;

# Classical multidimensional scaling
cmd <- (cmdscale(dis.ibd.matri))

# non-metric multidimensional scaling with SMACOF
smac <- mds(dis.ibd.matri, 14, type = "ordinal");

# compare both matrices
res.prot <- protest(smac$conf[,1:2], cbind(data.fibla$Lon_X, data.fibla$Lat_Y), permutations = 9999);

res.prot.cmd <- protest(cmd, cbind(data.fibla$Lon_X, data.fibla$Lat_Y), permutations = 9999);

clus <- Mclust(smac$conf[,1:5], G= 1:10)
plot(clus)

library("dbscan");
cl <- hdbscan(smac$conf[,1:2], minPts = 5)
plot(smac$conf[,1:2], col=cl$cluster+1, pch=20)
```


